Plug-and-play dual-tree algorithm runtime analysis

### 2014-11-26 bounds-files/sec01-supporting-lemmas.tex
Some changes and a better range search proof using Pari's assumptions.

We are somewhat limited in the sizes of $|S|$ we can scale to, because
calculating the expansion constant of a dataset is a computationally intensive
task.  Nonetheless, we can see that the expansion constant moves towards the
final value of $c_f$.  We postulate that the behavior of more complex real-world
datasets is similar.

>>>>

We are somewhat limited in the sizes of $|S|$ we can scale to, because
calculating the expansion constant of a dataset is a computationally intensive
task.  Nonetheless, we can see that the expansion constant moves towards the
final value of $c_f$, though potentially from above ({\it The results are not
done yet.  This paragraph may need rewriting.  But I strongly suspect the
results will show what I've said.}).  We postulate that the behavior of more
complex real-world datasets is similar.

### 2014-11-26 bounds-files/sec04-range-search.tex
Some changes and a better range search proof using Pari's assumptions.

    \IF{$d(p_q, p_r) \in [r_{\min}, r_{\max}]$ \AND \texttt{BaseCase($p_q$, $p_r$)} not yet called}
    \STATE  $N[p_q] \gets N[p_q] \cup \{ p_r \}$
    \ENDIF

@@

In the range search problem, the task is to find the set of reference points
%\begin{equation}
$N[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
%\end{equation}
for each query point $p_q$, where $[l, u]$ is the given
range.  The range count problem is practically identical, but only the size of
the set, $|N[p_q]|$, is desired.
Our proof works for both of these algorithms
identically, but we will focus on range search.  A \texttt{BaseCase()} and
\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
initialized to $\emptyset$ at the beginning of the traversal.

@@

\begin{equation}
\left| \{ p_r: p_r \in S_r, (1 - \alpha) l \le d(p_q, p_r) \le (1 + \alpha) u \}
\right| = O(1),
\end{equation}

\noindent the running time of range search or range count is bounded by
$O(c_r^{8 + \beta} (N + \theta))$ with $\theta$ defined as in Theorem
\ref{thm:ct-runtime} and
%\begin{equation}
$\beta = \log_2 (1 + \alpha^{-1})$.
%\end{equation}
\label{thm:rs}
\end{thm}

\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
$O(c_r^4 |R^*| (N + \theta))$. %  Thus, our only task is to bound the maximum size of the
%reference set, $|R^*|$.
By the pruning rule, for a query node
$\mathscr{N}_q$, the reference set $R^*$
is made up of reference nodes $\mathscr{N}_r$ that are within a margin of
$2^{s_q + 1} + 2^{s_r + 1} \le 2^{s_r^{\max} + 2}$ of the range $[l, u]$.  Given
that $p_r$ is the point in $\mathscr{N}_r$,

@@

\begin{equation}
B_{S_r}(p_q, u + 2^{s_r^{\max} + 1}) \setminus B_{S_r}(p_q, l - 2^{s_r^{\max} +
1}) \subseteq B_{S_r}(p_q, (1 + \alpha) u) \setminus B_{S_r}(p_q, (1 - \alpha)
l)
\end{equation}

\noindent and, due to the assumptions, $| B_{S_r}(p_q, (1 + \alpha) u) | - |
B_{S_r}(p_q, (1 - \alpha) l) | = O(1)$, $|R^*| = O(1)$.  Thus, the theorem
holds.
\end{proof}

@@

To understand the relation of the running time to the bounding quantity
$\alpha$, consider $\alpha =
(1 / 3)$; this yields a running time of $O(c^8 (N + \theta))$.  $\alpha = (1 /
7)$ yields $O(c^9 (N + \theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N +
\theta))$, and so forth.  As $\alpha$ gets smaller, the exponent on $c$ gets
larger, and diverges as $\alpha \to 0$.

The assumption that on number of points in the range of $[(1 - \alpha)l, (1 +
\alpha)u]$ for a single query is not too restrictive, but must be satisfied for
the result to be meaningful.  After all, if that assumption is not satisfied,
then it is not possible to perform range search in $O(1)$ time for each query
point, making an $O(N)$ bound for the whole query set impossible.

To show where the assumption is valid, consider that when presented with a very
large dataset, a practitioner may perform monochromatic range search on a range
$[l, u]$ for a geometric subset of their data; for instance, a hypersquare
with side length $2u$.  This might be done in order to find a range $[l, u]$
that produces a manageable number of results (if the range is too large, the
number of results returned for each query point may be enormous).  Then, that
practitioner may scale to a larger dataset by simply increasing the size of
their hypersquare.  In this setting, the number of points in the range of
$[(1 - \alpha)l, (1 + \alpha)u]$ for a single query remains $O(1)$ with respect
to $N$.  Alternately, it should be possible to tighten the range $[l, u]$ as $N$
increases in such a way that the number of points in the range remains constant.

@@

Alternately, if the assumption is not valid but it can be shown that the number
of points in the range $[(1 - \alpha)l, (1 + \alpha)u]$ is $O(k)$ where $k$ has
some dependence on $N$, then the running time can be easily bounded as $O(c^{4 +
\beta} k(N + \theta))$.

>>>>

    \IF{$d(p_q, p_r) \in [r_{\min}, r_{\max}]$ \AND \texttt{BaseCase($p_q$, $p_r$)} not yet called}
    \STATE  $S[p_q] \gets S[p_q] \cup \{ p_r \}$
    \ENDIF

@@

In the range search problem, the task is to find the set of reference points
%\begin{equation}
$S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
%\end{equation}
for each query point $p_q$, where $[l, u]$ is the given
range.  The range count problem is practically identical, but only the size of
the set, $|S[p_q]|$, is desired.
Our proof works for both of these algorithms
identically, but we will focus on range search.  A \texttt{BaseCase()} and
\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
initialized to $\emptyset$ at the beginning of the traversal.

In order to bound the running time of dual-tree range search, we require better
notions for understanding the difficulty of the problem.  Observe that if the
range is sufficiently large, then for every query point $p_q$, $S[p_q] = S_r$.
Clearly, for $S_q \sim S_r \sim O(N)$, this cannot be solved in anything less
than quadratic time simply due to the time required to fill each output array
$S[p_q]$.  Define the maximum result size for a given query set $S_q$, reference
set $S_r$, and range $[l, u]$ as

\begin{equation}
| S_{\max} | = \max_{p_q \in S_q} | S[p_q] |.
\label{eqn:smax}
\end{equation}

Small $| S_{\max} |$ implies an easy problem; large $| S_{\max} |$ implies a
difficult problem.  For bounding the running time of range search, we require
one more definition.

\begin{defn}
For a range search problem with query set $S_q$, reference set $S_r$, range $[l,
u]$, and results $S[p_q]$ for each query point $p_q$ given as

\begin{equation}
S[p_q] = \{ p_r : p_r \in S_r, l \le d(p_q, p_r) \le u \},
\end{equation}

\noindent define the {\it $\alpha$-expansion} of the range set $S[p_q]$ as the slightly larger set

\begin{equation}
S^{\alpha}[p_q] = \{ p_r : p_r \in S_r, (1 - \alpha) l \le d(p_q, p_r) \le (1 +
\alpha) u \}.
\end{equation}
\end{defn}

Using these notions, we may now bound the running time of range search.

@@

\begin{equation}
| S^{\alpha}[p_q] \setminus S[p_q] | \le C \; \; \forall \; p_q \in S_q,
\end{equation}

%\begin{equation}
%\left| \{ p_r: p_r \in S_r, (1 - \alpha) l \le d(p_q, p_r) \le (1 + \alpha) u \}
%\right| = O(1),
%\end{equation}

\noindent the running time of range search or range count is bounded by

\begin{equation}
O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N + \theta)
\right)
\end{equation}

with $\theta$ defined as in Theorem
\ref{thm:ct-runtime},
$\beta = \log_2 (1 + \alpha^{-1})$, and $S_{\max}$ as defined in Equation
\ref{eqn:smax}.
%\end{equation}
\label{thm:rs}
\end{thm}

\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
$O(c_r^4 |R^*| (N + \theta))$.  As with the previous proofs, then, our only task
is to bound the maximum size of the reference set, $|R^*|$.

By the pruning rule, for a query node $\mathscr{N}_q$, the reference set $R^*$
is made up of reference nodes $\mathscr{N}_r$ that are within a margin of
$2^{s_q + 1} + 2^{s_r + 1} \le 2^{s_r^{\max} + 2}$ of the range $[l, u]$.  Given
that $p_r$ is the point in $\mathscr{N}_r$,

@@

\begin{equation}
B_{S_r}(p_q, u + 2^{s_r^{\max} + 1}) \setminus B_{S_r}(p_q, l - 2^{s_r^{\max} +
1}) \subseteq B_{S_r}(p_q, (1 + \alpha) u) \setminus B_{S_r}(p_q, (1 - \alpha)
l).
\end{equation}

This set is necessarily a subset of $S^{\alpha}[p_q]$; by
definition, the number of points in this set is bounded above by $|S_{\max}| +
C$.  We may then conclude that $|R^*| \le |S_{\max}| + C$.  By taking the
maximum of the sizes of $|R^*|$ in both cases above, we obtain the statement of
the theorem.
\end{proof}

This bound displays both the expected dependence on $c_r$ and $|S_{\max}|$.  As
the largest range set $S_{\max}$ increases in size (with the worst case being
$S_{\max} \sim N$), the runtime degenerates to quadratic.  But for adequately
small $S_{\max}$ the runtime is instead dependent on $c_r$ and the parameter $C$
of the $\alpha$-expansion of $S_{\max}$.  This situation leads to a
simplification.

\begin{cor}
For sufficiently small $|S_{\max}|$ and sufficiently small $C$, the runtime of
range search under the conditions of Theorem \ref{thm:rs} simplifies to

\begin{equation}
O(c_r^{8 + \beta} (N + \theta)).
\end{equation}
\label{cor:rs}
\end{cor}

@@

In this setting we can more easily consider the relation of the running time to
$\alpha$.  Consider $\alpha = (1 / 3)$; this yields a running time of $O(c^8 (N
+ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + \theta))$, $\alpha = (1 /
15)$ yields $O(c^{10} (N + \theta))$, and so forth.  As $\alpha$ gets smaller,
the exponent on $c$ gets larger, and diverges as $\alpha \to 0$.

For reasonable runtime it is necessary that the $\alpha$-expansion of $S_{\max}$
be bounded.  This is because the dual-tree recursion must retain reference nodes
which may contain descendants in the range set $S[p_q]$ for some query $p_q$.
The parameter $C$ of the $\alpha$-expansion allows us to bound the number of
reference nodes of this type, and if $\alpha$ increases but $C$ remains small
enough that Corollary \ref{cor:rs} applies, then we are able to obtain tighter
running bounds.

%The assumption that on number of points in the range of $[(1 - \alpha)l, (1 +
%\alpha)u]$ for a single query is not too restrictive, but must be satisfied for
%the result to be meaningful.  After all, if that assumption is not satisfied,
%then it is not possible to perform range search in $O(1)$ time for each query
%point, making an $O(N)$ bound for the whole query set impossible.

%To show where the assumption is valid, consider that when presented with a very
%large dataset, a practitioner may perform monochromatic range search on a range
%$[l, u]$ for a geometric subset of their data; for instance, a hypersquare
%with side length $2u$.  This might be done in order to find a range $[l, u]$
%that produces a manageable number of results (if the range is too large, the
%number of results returned for each query point may be enormous).  Then, that
%practitioner may scale to a larger dataset by simply increasing the size of
%their hypersquare.  In this setting, the number of points in the range of
%$[(1 - \alpha)l, (1 + \alpha)u]$ for a single query remains $O(1)$ with respect
%to $N$.  Alternately, it should be possible to tighten the range $[l, u]$ as $N$
%increases in such a way that the number of points in the range remains constant.

@@

%Alternately, if the assumption is not valid but it can be shown that the number
%of points in the range $[(1 - \alpha)l, (1 + \alpha)u]$ is $O(k)$ where $k$ has
%some dependence on $N$, then the running time can be easily bounded as $O(c^{4 +
%\beta} k(N + \theta))$.
